### 2013-08-22 doc/dissertation/document/backmatter/mcsa_spn_estimator_analysis.tex
removing basic preconditioning sectiion and adding block jacobi subsection to the fuel assembly problem

\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
\label{subsec:spn_estimator_comparison}
The iterative performance of MCSA is studied in this appendix with
both the collision and expected value estimators using the ILUT
preconditioning. Although the expected value estimator achieved the
best iterative performance in analysis for the simple Poisson problem
in \S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective for
the $SP_N$ equations. The deterministic component of the expected
value estimator that couples the current state to other local states
through the iteration matrix stencil requires application of the
estimator to potentially several orders of magnitude more states at
every transition event.

>>>>

\chapter{MCSA Estimator and Fixed Point\\ Iteration Analysis}
\label{chap:spn_estimator_comparison}
The iterative performance of MCSA is studied in this appendix with
both the collision and expected value estimators using the ILUT
preconditioning. Although the expected value estimator achieved the
best iterative performance in analysis for the simple Poisson problem
in \S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective for
the $SP_N$ equations. The deterministic component of the expected
value estimator that couples the current state to other local states
through the iteration matrix stencil requires application of the
estimator to potentially several orders of magnitude more states at
every transition event.

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removing basic preconditioning sectiion and adding block jacobi subsection to the fuel assembly problem

%%---------------------------------------------------------------------------%%
\section{Basic Preconditioning for the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their diffusion-like form. In this section, we will
compute the spectral radius of the iteration matrix formed by the
$SP_N$ equations while parametrically varying the $SP_N$ order, the
$P_N$ order, the number of energy groups, and
upscattering/downscattering between energy groups for Jacobi
preconditioned problems to ensure that this type of preconditioning
will be applicable to the fuel assembly problem. For each problem in
the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{c}{Parameter}& 
      \multicolumn{1}{c}{Value} \\\hline
      Mesh Element Size & 1.0 \\
      Mesh Elements & $4 \times 4 \times 4$ \\
      Reflecting Boundaries & True \\
      Materials & 1 \\
      Source Strength & 1.0 \\
      Total Cross Section & 5.0 \\
      In-Group Cross Section & 0.25 \\
      Downscatter Cross Section & 1.0 \\
      Upscatter Cross Section & 0.1 \\
      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
      parameter study.} \textit{The parameters were fixed for each
      spatial location and energy group.}}
  \label{tab:spn_fixed_parameters}
\end{table}

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.

@@

Spectral radius computations for the block Jacobi preconditioned
iteration matrix were performed with Table~\ref{tab:group1bj} giving
the results for the 1-group case, Table~\ref{tab:group10dsbj} for the
10-group case with full downscatter only and
Table~\ref{tab:group10usbj} for the 10-group case with full
downscatter and full upscatter. A block size of $N_g\times(N+1)/2$ was
used for the preconditioner in all cases.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1bj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dsbj}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the block Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10usbj}
\end{table}

From the tabulated block Jacobi data it is clear that this is a viable
preconditioning choice for all $SP_N$ problems in terms of Monte Carlo
solution methods. All cases were observed to have a spectral radius
below unity. Based on these results, the block Jacobi method should be
the applicable to more complicated problems representative of physical
nuclear systems.

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
We first use block Jacobi preconditioning to solve the $17 \times 17$
fuel assembly problem. A single energy group problem was first
computed with $SP_1$ discretization, effectively giving the one-speed
neutron diffusion system for the fuel assembly resulting in 20,088
degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. To
investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Verification\ }
\label{sec:spn_mcsa_verification}
Through the numerical studies presented in this chapter, we have
developed a set of preconditioning techniques that permit MCSA to be
used with the $SP_N$ form of the neutron transport equation and
applied to a difficult fuel assembly criticality problem. Along with
these preconditioning techniques, a set a varying MCSA parameters were
analyzed to find those that yielded the best performance for this
particular problem. In addition, several steps were taken to mitigate
the dense composite linear operators that arise when performing
explicit MCSA preconditioning necessary for achieve convergence. Using
these results, we now compare MCSA to conventional Krylov solvers that
would be typically used to solve the $SP_N$ form of the transport
equation as presented here.

@@

All solvers will be preconditioned with ILUT using a drop tolerance of
\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
placed on the size of the subspace and therefore no restarts
occurred. When the collision estimator was used with MCSA, \sn{2}{4}
stochastic histories were used to compute the correction for every
energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
for the 2 and 4 group calculations respectively) corresponding to
approximately 1 stochastic history per DOF. When the expected value
estimator was used, \sn{5}{2} stochastic histories used for each
energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
4 group calculations) giving approximately 1 stochastic history for
every 5 DOFs. The reduced domain approximation was also applied in
conjunction with the ILUT preconditioning using a fill level of 100
and a threshold value of \sn{1}{-10} to reduce the density of states
in the Monte Carlo problem. For relaxation parameters, all MCSA
computations used a Neumann relaxation parameter of 0.7 while a
Richardson relaxation of 1.1 was used with the collision estimator and
1.0 with the expected value estimator. Additionally, both the
Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa}) and the
minimal residual-based MCSA iteration given by
Eq~(\ref{eq:mcsa_min_res}) will be
used. Table~\ref{tab:spn_solver_defs} gives definitions for the
solvers used to generate the results in the remainder of this
section. For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
group 0).
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Name}} & 
      \multicolumn{1}{l}{\textbf{Definition}} \\
      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
      GMRES-ILUT & GMRES preconditioned with ILUT \\
      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
                    & fixed point iteration and collision estimator \\
      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
                     & fixed point iteration and collision estimator \\
      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
                     & fixed point iteration and expected value estimator \\
      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
                      & fixed point iteration and expected value estimator \\
      Richardson-ILUT & Richardson iteration preconditioned with ILUT \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Solver definitions used for MCSA verification and
      performance analysis.} \textit{The ILUT preconditioner
      parameters were identical for all calculations and solvers.}}
  \label{tab:spn_solver_defs}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cl}\hline\hline
      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
      1 & \{ \sn{1}{-5} \} \\
      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Energy group lower bounds for the multigroup fuel
      assembly criticality problem in electron volts.} \textit{An
      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
  \label{tab:spn_group_structure}
\end{table}

>>>>

\subsection{Block Jacobi Preconditioning}
\label{subsec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their diffusion-like form. In this section, we will
the structure of the transport operator formed by the $SP_N$ equations
while parametrically varying the $SP_N$ order, the $P_N$ order, the
number of energy groups, and upscattering/downscattering between
energy groups. For each basic problem, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{c}{Parameter}& 
      \multicolumn{1}{c}{Value} \\\hline
      Mesh Element Size & 1.0 \\
      Mesh Elements & $2 \times 2 \times 2$ \\
      Reflecting Boundaries & True \\
      Materials & 1 \\
      Source Strength & 1.0 \\
      Total Cross Section & 5.0 \\
      In-Group Cross Section & 0.25 \\
      Downscatter Cross Section & 1.0 \\
      Upscatter Cross Section & 0.1 \\
      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
      parameter study.} \textit{The parameters were fixed for each
      spatial location and energy group.}}
  \label{tab:spn_fixed_parameters}
\end{table}

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively.

@@

\subsection{Block Jacobi Preconditioned Results}
\label{subsec:jacobi_prec_assembly_calc}
A single energy group problem was first computed with $SP_1$
discretization, effectively giving the one-speed neutron diffusion
system for the fuel assembly resulting in 20,088 degrees of freedom in
the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. To
investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Verification\ }
\label{sec:spn_mcsa_verification}
Through the numerical studies presented in this chapter, we have
developed a set of preconditioning techniques that permit MCSA to be
used with the $SP_N$ form of the neutron transport equation and
applied to a difficult fuel assembly criticality problem. In addition,
several steps were taken to mitigate the dense composite linear
operators that arise when performing explicit MCSA preconditioning
necessary for achieve convergence. Using these results, we now compare
MCSA to conventional Krylov solvers that would be typically used to
solve the $SP_N$ form of the transport equation as presented here.

@@

All solvers will be preconditioned with ILUT using a drop tolerance of
\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
placed on the size of the subspace and therefore no restarts
occurred. When the collision estimator was used with MCSA, \sn{2}{4}
stochastic histories were used to compute the correction for every
energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
for the 2 and 4 group calculations respectively) corresponding to
approximately 1 stochastic history per DOF. When the expected value
estimator was used, \sn{5}{2} stochastic histories used for each
energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
4 group calculations) giving approximately 1 stochastic history for
every 5 DOFs. The reduced domain approximation was also applied in
conjunction with the ILUT preconditioning using a fill level of 100
and a threshold value of \sn{1}{-10} to reduce the density of states
in the Monte Carlo problem. For relaxation parameters, all MCSA
computations used a Neumann relaxation parameter of 0.7 while a
Richardson relaxation of 1.1 was used with the collision estimator and
1.0 with the expected value estimator. Additionally, both the
Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa}) and the
minimal residual-based MCSA iteration given by
Eq~(\ref{eq:mcsa_min_res}) will be used. A supplementary performance
analysis for this set of MCSA variations is presented in
Appendix~\ref{chap:spn_estimator_comparison}.

Table~\ref{tab:spn_solver_defs} gives definitions for the
solvers used to generate the results in the remainder of this
section. For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
group 0).

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Name}} & 
      \multicolumn{1}{l}{\textbf{Definition}} \\
      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
      GMRES-ILUT & GMRES preconditioned with ILUT \\
      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
                    & fixed point iteration and collision estimator \\
      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
                     & fixed point iteration and collision estimator \\
      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
                     & fixed point iteration and expected value estimator \\
      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
                      & fixed point iteration and expected value estimator \\
      Richardson-ILUT & Richardson iteration preconditioned with ILUT \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Solver definitions used for MCSA verification and
      performance analysis.} \textit{The ILUT preconditioner
      parameters were identical for all calculations and solvers.}}
  \label{tab:spn_solver_defs}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cl}\hline\hline
      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
      1 & \{ \sn{1}{-5} \} \\
      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Energy group lower bounds for the multigroup fuel
      assembly criticality problem in electron volts.} \textit{An
      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
  \label{tab:spn_group_structure}
\end{table}
